Gas exchange in large rivers influenced by largest turbulent eddies: implications for remotely sensing gas exchange via the SWOT satellite

### 2021-09-30 manuscript_v5.Rmd
model update to the chainsaw model

---
title: "Gas exchange in large rivers influenced by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document: 
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points
- River channel efficiency effects the relationship between gas exchange and dissipation of turbulence
- Wide, less "efficient" river channels exert greater influence on near-surface turbulence than narrower, more efficient ones
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data

@@

## Abstract
*AGU Advances. So 250 words here* <br>
*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style.*

@@

## 1 Introduction
Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used in this study.

@@

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. More broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020]. Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010]. This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity and *Sc* is the Schmidt number. Note that we directly measuring $\epsilon_S$ is non-trivial and though a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface, this necessiarly assumes that bed shear alone effects turbulence which is often not the case.
$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$

While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear or wind shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems. First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a]. Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc. further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999]. Those authors proposed modeling fluvial *k* via form-drag dissipation $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation. Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it should be stressed that $\epsilon_S \neq \epsilon_D$. This is because $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate. Consequently, the, $k\sim\epsilon_D$ relationship usually has a slope much steeper than the 1/4 in equation 2.

Compared to small streams, relatively little attention has been paid to gas exchange in very large rivers. Large rivers are often conceptualized as a hybrid condition of both fluvial and non-fluvial *k* dynamics. To date, the handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021]. This is also presumably limited to forested environments that provide shelter from wind.

These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets. They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers. However, they did not directly compare against *k* measurements made in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for their results.

Finally, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a]. It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020]. This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020]. Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].

In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change along the stream-to-river continumn and how does this relate to one of the classical models of gas evasion in aquatic systems (equation 2)? We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations. The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

@@

![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced. Not mapped here are over 170,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al. 2019) and over 700 gas exchange velocity measurements (Ulseth et al. 2019).](cache/maps/map_fin.jpg)

Gas exchange theory (section 3) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics. Ultimately, this dataset was filtered down to 171,553 discrete measurements for this study (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on 47 SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4. These are two models for the Saint Lawrence River and one for the Tanana River.

To assess the influence of measurement error on BIKER's performance (section 4), we use the error model developed by @durandHowWillRadar2020a and implemented on 17/47 of the rivers by @frassonExploringFactorsControlling2021. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020a]. Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.

For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

## 3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century. As noted in the introduction, turbulence in river flows is usually generated by shear forces at the riverbed, but also across the entire flow depth via 'form-drag' from riverbanks, meanders, bars, suspended sediment, etc [@moogStreamReaerationNonuniform1999]. This concept is synonymous with the geomorphic idea of available stream power, or the total energy dissipated from a river reach via the liquid's potential energy along its slope [originally derived from first-principles by @bagnoldSedimentDischargeStream1960; @bagnoldApproachSedimentTransport1966]. Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ (simply the available stream power normalized by the water mass) is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.

While form-drag dissipation is effectively an alternative representation of reach-scale stream power, little attention has been paid to the resulting scale-dependecy of $\epsilon_D$. Available stream power, and thus 'form-drag' shear, exists at the depth-scale and is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999]. This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, its depth will decrease and its wetted perimeter (and therefore total frictional resistance) will increase. This concept is illustrated graphically in Figure 2 and relates the geomorphic concept of 'river channel efficiency' to turbulence: all else being constant, a less efficient channel (wide/shallow) will have greater frictional resistance and therefore form-drag (Chow, 1959). As shown via the orange arrows in Figure 2, the inefficient channel should 1) produce more turbulence, and 2) turbulence generated via form-drag should more readily reach the surface (the red arrows in Figure 2). Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$ (channel area divided by the width). An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.

![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate. All else being constant, a less efficient channel shape (wide/shallow) should 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag should more readily reach the surface. This is illustrated by the orange arrows, where the inefficient channel arrows more frequently reach the surface (and influence surface turbulence- red arrows).](cache/k600_theory/conceptual_model.jpg)

>>>>

---
title: "Gas exchange in large rivers influenced by largest turbulent eddies: implications for remotely sensing gas exchange via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Key Points
- Both large and small scale eddies control gas exchange in hydraulically wide rivers
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data
- BIKER is also robust to measurement errors internal to the SWOT river measurements

@@

## Abstract
UPDATE

Extensive research over the past two decades has shown that the global river network emits significant amounts of carbon dioxide ($CO_2$) to the atmosphere via gas evasion, a diffusive flux across the air-water interface. This global flux is still relatively poorly constrained, with much ambiguity coming from mechanistic and predictive uncertainties in the gas exchange velocity (the rate of this diffusion). We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of river hydrology research at the global scale, will provide similarly transformative insights to fluvial gas exchange modeling. Here, we present work predicting gas exchange velocity using SWOT observations. We exploit the unique geomorphology of SWOT-observable rivers to develop a process-based model that is nearly entirely remotely sensible. We then couple this model with established Bayesian techniques to develop BIKER, or the 'Bayesian Inference/Inversion of the k600 Exchange Rat' algorithm. We validate BIKER on 47 SWOT-simulated rivers and over 700 discrete measurements of river gas exchange, yielding 1) a theoretical model that explains 64% of variation in gas exchange in SWOT-observable rivers and 2) an algorithm that predicts this theoretical model solely from SWOT observations with a median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$nrmse), 2)`. Additionally, by-river performance improves as the amount of available SWOT data increases. BIKER is also robust to measurement errors implicit in the SWOT data, with a median normalize RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==1,]$nrmse), 2)`. With SWOT set to launch in 2022, we suggest that BIKER might address the lack of data on gas exchange in large rivers and help to better constrain current and future fluvial gas evasion models at the global scale.

@@

## 1 Introduction
Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used in this study.

@@

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. More broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009 for a review] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020 for a review]. Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010]. This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, *Sc* is the Schmidt number. Note that directly measuring $\epsilon_S$ is non-trivial and a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface (Appendix A).

$$\mathbf{(2)} k\sim Sc^{-1/2}(v\epsilon_S)^{1/4}$$

While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear or wind shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems. First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019]. Another approach is to model *k* via 'form-drag shear' (i.e. total stream power per unit mass water) that captures frictional resistance via river channel banks, meanders, bars, etc. [@moogStreamReaerationNonuniform1999]. Authors have since shown that this model reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019], though it requires good estimates of the average flow velocity. Compared to small streams, relatively little attention has been paid to gas exchange in very large rivers. Large rivers are often conceptualized as a hybrid condition of both fluvial and non-fluvial *k* dynamics. The handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021]. These mechanistic uncertainties are additionally limited by a dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*. They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers. However, their simulated rivers were limited by a lack of hydraulic measurements and had to rely on hydraulic geomtery models to estimate river depth and velocity. Further, they did not assess model fit in very steep systems [@ulsethDistinctAirWater2019] and stopped short of parsing out hydraulic explanations for their results.

These mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953]. It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2019].

A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2019]. This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2019]. Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].

In this context, here we borrow tools from fluvial geomorphology and revist the fundamental mechanisms behind gas exchange in large rivers to answer the following question: does the small-eddy model work in large rivers like those that SWOT will observe? We exploit the findings from this analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations. The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

@@

![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced. Not mapped here are over 170,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al. 2019) and 763 gas exchange velocity measurements (Ulseth et al. 2019; Raymond et al. 2012; Churchill et al. 1964; Owens et al. 1964).](cache/maps/map_fin.jpg)

Gas exchange theory (section 3) is explored using multiple datasets which measured and/or collected from the literature 763 measurements of stream hydraulics and $k_{600}$. This is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. 701 of these measurements were previously collected from other studies by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements were added by us but measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics. Ultimately, this dataset was filtered down to 171,553 discrete measurements for this study (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on 47 SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4. These are two models for the Saint Lawrence River and one for the Tanana River.

To assess the influence of measurement error on BIKER's performance (section 4), we use the error model developed by @durandHowWillRadar2020 and implemented on 17/47 of the rivers by @frassonExploringFactorsControlling2021. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020]. Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.

For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

## 3 Hydraulic geometry in large rivers and implications for scaling fluvial gas exchange velocity
Rivers and streams change predictably along their longitudinal profile from headwater to ocean. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth (i.e. $R_h \approx H$). This is a common assumption in hydraulic and geomorphic modeling of large rivers, and occurs in `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of our SWOT-observable river measurements but only `r round(percs_overall[percs_overall$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of all `r sum(percs$n)` river measurements (see Text S1 for how we arrived at these numbers). These channels are often referred to as 'hydraulically wide'.

To model gas exchange velocity in hydraulically wide channels, we make some simplifying assumptions to equation 2. First, we impose the log-law-of-the-wall model for $\epsilon_S$ (Appendix A) on to the small-eddy model (equation 2) and simplify $R_H = H$, arriving at equation 3. Equation 4 further simplifies this for *k* normalized to a Schmidt number of 600, and with statistical parameter $\alpha$ reflecting both the scaling as well as the constant Schmidt number and viscosity [@wangPhysicallyBasedScaling2021]. Equation 4 thus defines gas exchange velocity solely as a function of water surface slope and average flow depth without a reliance on shear velocity, mean flow velocity, or $\epsilon_S$ at all. Again, this is hypothetically only valid in a wide channel.

$$\mathbf{(3)} k \sim Sc^{-1/2}v^{1/4}(gS)^{3/8}H^{1/8}$$
$$\mathbf{(4)} k_{600}=\alpha(gS)^{3/8}H^{1/8}$$

Alternatively, @moogAirWaterGasTransfer1999 noted that low-turbulence uniform channel flows often do not exhibit homogenous surface dissipation at their air-water interface. They proposed a conceptual 'chainsaw model' that relies on both small-scale eddies and large, depth-scale ones. It limits surface dissipation to only 'active zones' within which a homogenous dissipation rate occurs and the small-eddy model is valid. But, the fraction of the interface that are active zones is proportional to the larger, depth-scale eddies. This means that if the entire interface is treated as an active zone, the model would simply reduce to the small-eddy model. We chose to also test this model given that SWOT-observable rivers are generally large and low-turbulence (here, low-turbulence is realitive to the highly-turbulent water surfaces of small, steep streams with significant whitewater). Further, to our knowledge the chainsaw model has never been empirically tested, and so we explore it here. To do this, we impose a hydraulically wide channel on the chainsaw model just like we did to the small-eddy model and arrive at equation 5 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 4), except for the exponents, which are both steeper.

$$\mathbf{(5)} k_{600}=\beta(gS)^{9/16}H^{11/16}$$

We test the performances of equations 4 and 5 on our dataset of field-measured $k_{600}$ in Figure 3. Figure 3a fits the hydraulically-wide small-eddy model (equation 4) to hydraulically-wide measurements: only the data where $R_h$ is within 1% percent of $H$. Figure 3b fits the hydraulically wide chainsaw model (equation 5) to the same subset of data.

![Figure 3: Model development for predicting gas exchange velocity from SWOT-observable variables. (a) Empirical confirmation of the hydraulically wide small-eddy model. (b) Empirical confirmation of the hydraulically-wide chainsaw model. (c) Cummulative density function (CDF) of the hydraulic radius/mean flow depth ratios for all SWOT-observable measurements. Nearly all of these measurements would be considered hydraulically wide.](cache/k600_theory/k600Plot.jpg)
